Overview and Status of Assumptions of Physics\footnote{http://assumptionsofphysics.org} \\ \ifextended (Extended version) \else (Short version) \fi

### 2019-12-28 Articles/Overview/overview.tex
Preparing for first draft release

\newcommand\fact[2]{\ifextended \emph{#1}.#2 \fi}

@@

\subsection{Index}
To this paper, not to the work.

- Goals and method
Goals: Understanding, have a better sense of the math, give a background for new theories...
Methods: reverse engineering vs forward engineering, holistic

- General theory
relationship between math and physics
logic of verifiable statements
properties and quantities
accuracy
differentiability

- Fundamental physics
states and processes
determinism/reversibility: isomorphism in the category
entropy

- Physical theories
Classical mechanics
Quantum mechanics
Thermodynamics
Statistical mechanics

@@

As each topic is investigated by the project, it generally goes through three phases. We mention them here as they will give context to the status of the project and to people who may want to contribute.

\textbf{Reverse engineering.} The goal of this phase is to deconstruct an already existing theory or mathematical structure to understand what each piece is supposed to represent physically. One way to conduct such investigation is to ask what would happen if a particular characteristic of the framework would fail.\footnote{For example, a line of reasoning could be as follows: Hamilton's equations are differential; if they weren't the mapping of densities and areas would not be well defined; maybe Hamiltonian mechanics is not about mapping points but about mapping densities or areas.} This usually leads to a set of necessary assumptions about the physical system one is describing, the ones one must take for the particular description to apply.

\textbf{Forward engineering.} The goal of this phase is to construct an argument from the necessary assumptions outlined in the previous phase to rederive the framework at hand. This checks whether the necessary assumptions are also sufficient \footnote{For example, if we started simply by assuming we have density distributions mapped in time, this would give us all possible differentiable evolutions, not necessarily Hamiltonian ones.}. It ends when a set of necessary and sufficient conditions to rederive the theory are found, meaning that the full theory has been physically characterized.

\textbf{Formalization.} The goal of this phase is to take the previous set of arguments and formalize them in pure mathematical language. This forces us to clarify all the starting points and what parts of the arguments are truly formal and what are physical. Unfortunately, this can ultimately be done only by starting from the very foundations\footnote{For example, before formalizing Hamiltonian mechanics we have to formalize differentiable manifolds, real numbers and topological spaces.} though we are at a point where an overall tentative structure is emerging and tentative formalization can be attempted on higher structures.

@@

@@

@@

@@

@@

The general theory at this point covers only the topological aspects and needs to be extended to the geometrical/measure theoretical ones. The idea should be that diverse aspects, such as geometry, measure theory, probability theory, information theory, can be recovered by characterizing the granularity (i.e. accuracy) of the statements. As each statement corresponds to a Borel set of possibilities, assigning a size to these sets effectively means giving a size to the statements themselves. For geometry, two sets have the same volume because they contain the same number of possible spatial points. For probability, borrowing idea from Cox probable inference, we would assign probability only on conditional statements, therefore the $P(s_1 | s_2)$ would measure what fraction of the possibilities compatible with $s_2$ are also compatible with $s_1$. For information theory, given two statements the logarithm of the ratio of the size of the possibilities corresponds to the number of yes/no questions that are needed to go from one statement to the other, the relative information.

These ideas still need to be developed in a full fledged theory to see whether they can work all the way through.

@@

@@

@@

\subsection{Other theories}

@@

We haven't yet attempted to work with general relativity as of yet. Classical mechanics as derived in this project is already relativistic, in the sense that we derive a metric tensor that is locally Minkowskian. However, it is not clear whether that metric has a link to the energy/momentum tensor or it is independent.

>>>>

\newcommand\fact[2]{\ifextended \emph{#1}. #2 \fi}

@@

@@

As each topic is investigated by the project, it generally goes through three phases. Reverse engineering, where we try to find what assumptions are beneath a theory. Forward engineering, where we try to check whether the theory can be recovered from the assumptions we found. Formalization, where we try to codify in mathematical language the assumptions and the derivation.

\fact{Reverse engineering} {The goal of this phase is to deconstruct an already existing theory or mathematical structure to understand what each piece is supposed to represent physically. One way to conduct such investigation is to ask what would happen if a particular characteristic of the framework would fail.\footnote{For example, a line of reasoning could be as follows: Hamilton's equations are differential; if they weren't the mapping of densities and areas would not be well defined; maybe Hamiltonian mechanics is not about mapping points but about mapping densities or areas.} This usually leads to a set of necessary assumptions about the physical system one is describing, the ones one must take for the particular description to apply.}

\fact{Forward engineering} {The goal of this phase is to construct an argument from the necessary assumptions outlined in the previous phase to rederive the framework at hand. This checks whether the necessary assumptions are also sufficient \footnote{For example, if we started simply by assuming we have density distributions mapped in time, this would give us all possible differentiable evolutions, not necessarily Hamiltonian ones.}. It ends when a set of necessary and sufficient conditions to rederive the theory are found, meaning that the full theory has been physically characterized.}

\fact{Formalization} {The goal of this phase is to take the previous set of arguments and formalize them in pure mathematical language. This forces us to clarify all the starting points and what parts of the arguments are truly formal and what are physical. Unfortunately, this can ultimately be done only by starting from the very foundations\footnote{For example, before formalizing Hamiltonian mechanics we have to formalize differentiable manifolds, real numbers and topological spaces.} though we are at a point where an overall tentative structure is emerging and tentative formalization can be attempted on higher structures.}

@@

\fact{Different algebra}{Verifiable statements do not follow the standard Boolean algebra. The negation of a verifiable statement is not necessarily verifiable since the test is not guaranteed to terminate. The conjunction of a finite set of verifiable statements is necessarily verifiable since we can run the tests one by one and, if all are successful, the conjunction is also verified. This cannot be extended to an infinite set since the test would not terminate. The disjunction of a countable set of verifiable statements is necessarily verifiable since, once one test terminates, the disjunction is verified and we can stop. This cannot be extended to an infinite set that is more than countable since we would not be able to find the test that terminates in a finite time.}

\fact{Decidable statements}{These are statements that can be verified and falsified. That is, their test always terminates. These can be defined from verifiable statements and they follow a finite Boolean algebra: the negation, finite conjunction and finite disjunction of decidable statements is also decidable.}

@@

\fact{Maximum cardinality for the possibilities}{Because the domain is generated by a countable set of verifiable statements, the possibilities must be distinguished with a countable sequence of boolean values. This means that the cardinality of the possibilities cannot exceed that of the continuum. This means that higher order infinities and all associated problems do not play a role in science.}

@@

\fact{Interior, exterior and boundary}{Given a test associated to a theoretical statement, it will either terminate successfully, terminate unsuccessfully or never terminate. Each Borel set is associated with a theoretical statement, which means it is associated with a test. The interior of the Borel set represents the cases in which the test will terminate successfully; the exterior represents the cases in which the tests will terminate unsuccessfully; the boundary represents the cases in which the test will not terminate.}

\fact{Decidable and undecidable statements}{A theoretical statement is decidable if the associated test always terminates, which means the boundary is empty. This corresponds to a clopen set. It is undecidable if the associated test never terminates, which means the boundary is the full set.\footnote{For example, ``there are three apples in the bowl'' is a decidable statement, while ``the mass of the electron is a rational number when expressed in eV'' is undecidable.}}

@@

\fact{Continuity and inference}{An inference relationship establishes that testing a verifiable statement in one domain tells whether a statement in the other is true or false. As these statements are equivalent, the experimental test on one constitutes an experimental test on the other, which means the second statement is verifiable. One cannot impose an inference relationship between a verifiable and non-verifiable statement. In terms of the natural topology, these correspond to open sets, therefore, by definition, the map is continuous as it maps open sets to open sets. The typical requirement that functions are ``well behaved'', then, is actually physically significant.}

@@

The general theory at this point covers only the topological aspects and needs to be extended to the geometrical/measure theoretical ones. Our current thinking is that diverse aspects, such as geometry, measure theory, probability theory, information theory, can be recovered by characterizing the granularity (i.e. accuracy) of the statements. Group theoretical aspects may be recovered

\fact{Granularity as foundational concept} {As each statement corresponds to a Borel set of possibilities, assigning a size to these sets effectively means giving a size to the statements themselves. For geometry, two sets have the same volume because they contain the same number of possible spatial points. For probability, borrowing idea from Cox probable inference, we would assign probability only on conditional statements, therefore the $P(s_1 | s_2)$ would measure what fraction of the possibilities compatible with $s_2$ are also compatible with $s_1$. For information theory, given two statements the logarithm of the ratio of the size of the possibilities corresponds to the number of yes/no questions that are needed to go from one statement to the other, the relative information.}

\fact{Recover differential geometry or geometry measure theory} {The idea is to start with finite functions of finite k-dimensional surfaces that have the following properties: they are linear if the surface is broken into disjoint parts (the value for the whole is the sum of the values for the parts) and they respect the limit (the value for the limit surface is the limit of the sequence of values for the surfaces). Under these, and possibly others, conditions we expect these functions to be expressed through an integral of local functions (i.e. k-forms) over infinitesimal surfaces (i.e. k-vectors).}

@@

\fact{Determinism and reversibility is isomorphism}{TODO}

@@

\fact{Link between quantum hydrodynamics, stochastic mechanics, Bohmian mechanics}{We should be able to reconcile the fact that quantum mechanics can be reframed in these different ways. The fact that hydrodynamics-like equations emerge should be linked to the idea that a quantum is a finite size object that can be thought as made of infinitesimal parts, as a fluid. Each part, though, is subjected by the Brownian motion associated with the unknowable internal dynamics. Stochastic mechanics, then, characterizes those fluctuations while Bohmian mechanics characterizes the average paths.}

\fact{Deriving quantum from classical}{If the above characterizations are understood, we should be able, in line of principle, to take a classical system, couple it with a suitable purely stochastic dynamic and recover quantum mechanics.}

@@

The work still needs to be extended to field theories, classical (electromagnetism and general relativity) and quantum (quantum electrodynamics, quantum chromodynamics, ...). Some ideas have been explored but only at a preliminary stage.

@@

\fact{General relativity} {Classical mechanics as derived in this project is already relativistic, in the sense that we derive a metric tensor that is locally Minkowskian. However, it is not clear whether that metric has a link to the energy/momentum tensor or it is independent. One idea to explore is that the presence of a field changes how states are counted over kinetic momentum (i.e. $\{mu^\alpha, mu^\beta\} = F^{\alpha\beta}$). It remains to be seen whether the curvature of the metric can be understood in terms of this insight.}
